15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
figure 6 fixes & edits to discussion

---
title: '15yrsChange'
output: 
    #pdf_document: 
    bookdown::word_document2: 
      md_extensions: +raw_html-markdown_in_html_blocks
      pandoc_args: ['--lua-filter', 'read_html.lua']
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: citations.bib
csl: new-phytologist.csl

@@

![**_Figure `r recruitment_by_genera`. Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across areas differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment > 20x that of canopy species recruitment.](display/Figure5.jpeg)

@@

Despite inherent limitations in extrapolating findings into the future and from a single site to a broader region, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threat_2020] on forests across the region.
<!--MOVED UP:
The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020]. -->
Also paralleling our site, persistently elevated deer populations are reported throughout eastern US forests, and pressures from deer brousing have contributed to regneration debts of canopy species throughout the region [@miller_compounding_2019; @miller_overabundant_2023].

Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forest ecosystem processes.
Nuisance species, driven by human introduction and perturbations to ecological equilibria, interact with other anthropogenic changes to modify forest ecosystems.
By increasing mortality, nuisance species have the capacity to alter short-term carbon dynamics, resulting in large carbon losses that would not be expected in models that exclude them.
Through recruitment suppression, nuisance species can have long-term effects that may alter successional dynamics. They can delay recovery from disturbances, anthropogenic or natural, and even suppress regeneration into a closed-canopy for sustained periods.
They also can decrease resilience to other pressures as pests and pathogens decrease canopy diversity by shrinking the number of possible canopy species.

Nuisance species pose a serious threat to the health of temperate forests.
But, this work highlights the potential of active management to protect them.
Increased understory recruitment within the deer exclosure demonstrates that lowering deer populations can reduce browsing pressure.
The direct linkage between non-indigenous pests and pathogens and decreased carbon storage demonstrates the need for natural climate solutions that focus on preserving the forests that we have instead of planting new ones.
Opportunities exist to expand efforts that prevent further spread, increase the application of extant treatments, and continue research into natural immunity.

<!--# Future Directions

Understanding interactions between global change drivers, an ongoing area of research, is critical. We would expect that trees unaffected by nuisance species (*e.g.* *Liriodendron tulipifera*, *Quercus rubra*) might see productivity increases as canopy gaps form and competition from nearby trees decreases. However, continuous decreases in AWG throughout the plot suggests that other mechanisms, including increasing maximum temperatures during the growing season (Dow et al 2022), may be hindering tree growth. Tree seedlings of many species, already decimated by deer browsing, are further threatened by rising maximum temperatures (McEwan et al 2011). Increasing mortality within *Quercus* species is another focal area of future research, as species within the *Quercus* genera constitute > 33% of woody productivity at this site (McGregor et al 2020). Multiple global changes already impact the SCBI ForestGEO plot; carbon storage in the site is, and will be, mediated by multifaceted, interacting pressures resulting from human actions.

Future research should build on our results by expanding these analyses to other temperate forest regions. While other areas will have different forest compositions and herbivory dynamics, increasing pressure from pests and pathogens is pervasive throughout temperate forests (Santini et al. 2012, Ayres and Lombardero 2018). Moreover, while not a large factor at this site, the spread of non-indigenous plant species can further threaten the  regeneration of woody canopy species (Pile Knapp et al, 2023). Large forest dynamics sites (*i.e.* other sites in the ForestGEO network) are well-suited to this research and can be leveraged to quantify how nuisance species compromise present and future forest carbon sequestration. Expanding the scale of these analyses may also facilitate the incorporation of nuisance species into global vegetation and carbon models. Future evaluations of temperate forest health and carbon storage must account for the substantial effects of nuisance species along with other global changes. -->

>>>>

---
output: 
    #pdf_document: 
    bookdown::word_document2: 
      md_extensions: +raw_html-markdown_in_html_blocks
      pandoc_args: ['--lua-filter', 'read_html.lua']
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: citations.bib
csl: new-phytologist.csl

@@

![**_Figure `r recruitment_by_genera`. Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across areas differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment > 20 times that of canopy species recruitment.](display/Figure5.jpeg)

@@

Despite inherent limitations in extrapolating findings into the future and from a single site to a broader region, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threat_2020] on forests across the region.
Also paralleling our site, persistently elevated deer populations are reported throughout eastern US forests, and pressures from deer brousing have contributed to regneration debts of canopy species throughout the region [@miller_compounding_2019; @miller_overabundant_2023].

Accounting for nuisance species in carbon models and future forest predictions is imperative. Our results underscore how ignoring nuisance species at our site, and likely much of the region, creates an expectation that temperate decidious forests will keep growing steadily and accumulating C [@wu_uncertainty_2023]. Instead, by increasing tree mortality, nusicance species alter short-term C dynamics and result in large decreases in living woody biomass that are not predicted by models that fail to account for their impact. A result of human introduction and perturbations to ecological equilibria, nuisance species co-occur with other anthropogenic changes and may have interactive effects on forest ecosystems. Through recruitment suppression, nuisance species can delay recovery from disturbances, anthropogenic or natural [@bradshaw_impacts_2016; @harvey_emergent_2023], and may even suppress regeneration into a closed-canopy for sustained periods. Unlike many other disturbances, many non-indigenous pests and pathogens not only induce mortality, but then prevent recovery of their host species [@conrad_threat_2020; @reed_distribution_2022; @knight_factors_2013]. Elimination of canopy species reduces biodiversity, which can decrease resilience to other disturbances (*e.g.* drought; [@anderegg_hydraulic_2018]) further endagerning forest health and C sequestration potential.

Nuisance species pose a serious threat to the health of temperate forests. However, active forest management offers pathways to protect these ecosystems.
At our site, increased understory recruitment within the deer exclosure supports  that lowering deer populations can reduce browsing pressure. 
Interventions that help reduce, or even eliminate, pests and pathogens  (*e.g.* monitoring programs, eradication technologies) exist and are improving  [@simberloff_impacts_2013]. Management of nuisance species at the scales necessary unquestionably will require enormous investment and effort. However, protecting existing forests represents a far more effective natural climate than emerging efforts to plant new ones [@cook-patton_protect_2021].
